%-----------------------------------------------------------------------------
%  Bernoulli-EKF Multi-Target Tracker on SE(3) — design + implementation
%-----------------------------------------------------------------------------
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools,bm,enumitem,hyperref}
\usepackage{algpseudocode,algorithm}

\DeclareMathOperator{\Tr}{tr}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\Log}{Log}
\newcommand{\R}{\mathbb{R}}
\newcommand{\SE}{\mathrm{SE}(3)}
\newcommand{\so}{\mathrm{SO}(3)}
\newcommand{\se}{\mathfrak{se}(3)}
\newcommand{\Ad}{\mathrm{Ad}}
\newcommand{\file}[1]{\texttt{#1}}

\title{Bernoulli-EKF on $\SE$:\\
  A Compact Multi-Object Tracker for Mobile Manipulation}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Part I — Algorithmic Overview}
\label{sec:overview}

The system tracks per-object Bernoulli-Gaussian beliefs over rigid
body poses on $\SE$, conditioned on RGB-D detections from a moving
camera mounted on a manipulator. Per frame $k$ it ingests the SLAM
base-pose $T_{wb,k}$, the camera-to-base extrinsic $T_{bc,k}$, the
gripper-to-base $T_{bg,k}$, the gripper finger width, and a list of
SAM2 detections $\{(\mathrm{pid}_d, \mathrm{label}_d, \mathrm{mask}_d,
\mathrm{score}_d)\}_d$ projected through depth.

\paragraph{1. State.} Each track $o$ carries
$(\mu_{bo}, \Sigma_{bo}, r_o, \ell_o, \tau_o)$:
$\mu_{bo} \in \SE$ is the object pose in the \emph{base} frame,
$\Sigma_{bo} \in \R^{6\times 6}$ is its covariance in the body-frame
tangent at $\mu_{bo}$, $r_o \in [0,1]$ is the existence probability,
$\ell_o$ a label histogram, and $\tau_o$ the SAM2 perception id.
\emph{Base-frame storage is the canonical choice}: SLAM uncertainty
$\Sigma_{wb}$ never enters the recursion; the world view is the
deterministic composition $T_{wo} = T_{wb}\,\mu_{bo}$.

\paragraph{2. Predict.} Two motion models, applied per-track:
world-static (default, $u_k = T_{wb,k}^{-1}T_{wb,k-1}$, applied via
the adjoint to $\Sigma$ plus per-phase Q) or rigid-attached
(held set, $\delta_{bg,k} = T_{bg,k}T_{bg,k-1}^{-1}$).

\paragraph{3. Detection conditioning.} Subpart suppression
(\file{det\_dedup.py}) absorbs detections whose 3D voxel sets are
mostly contained in another's; centroid back-projection drops masks
without valid depth.

\paragraph{4. Association.} Translation-only Mahalanobis with a
$\chi^2_3$ gate, $G_{\text{out},t}=21.108$, $R_{cam}=(0.02\,m)^2 I$.
Hungarian on the gated cost; SAM2-id continuity bonus $\alpha=4.4$
breaks ties.

\paragraph{5. Update.} For matched pairs on the held track, two
safety prefilters guard the Joseph step:
held-prefilter (centroid-to-$T_{we}$ within 0.25~m) and
innovation-clamp (centroid-to-$\mu_w$ within 0.20~m). The Joseph step
runs on the \emph{translation block only}; the rotation block and
cross-cov are left zeroed (a translation-only observation carries no
rotation information). This keeps $\Sigma_{bo}$ PSD across long runs.

\paragraph{6. Birth and prune.} A confirm-3 / score-0.30 / border /
near-live (default 5~cm; 25~cm against $T_{we}$ for the held oid)
chain controls births. Prune when $r<0.05$, except hold the held
oid above $r=0.5$.

\paragraph{7. Self-merge.} Same-label tracks within 5~cm fuse via
information sum, EXCEPT pairs in the relation-protected set
(scene graph asserts they are distinct). The held seed is always
the keeper.

\paragraph{8. Relations.} A VLM/REST backend
(\file{relation\_orchestrator.py}) is fired at first call,
release transitions, new-oid arrivals, and every 90 frames. Its
parent matrix is converted to \texttt{RelationEdge} records with the
convention \textbf{parent = supported, child = supporter} (LLM's $i$
is the supporter; we emit $(j,i,\text{on})$). A per-edge EMA
($\alpha=0.3$, threshold $0.5$) smooths flicker.

\paragraph{9. Held-set expansion.} The set carried by the gripper is
the closure of $\{\text{held seed}\}$ under in/on edges; after each
self-merge, EMA keys are remapped (drop $\to$ keep) so the closure
stays valid.

\bigskip\hrule\bigskip

\section*{Part II — Detailed Elaboration}
\label{sec:detail}

\subsection*{II.1 State}

\file{pose\_update/gaussian\_state.py}.
$$\mathcal{S}_o = (\mu_{bo}\in\SE,\ \Sigma_{bo}\in\R^{6\times 6}_{\succeq 0},\
  r_o\in[0,1],\ \ell_o,\ \tau_o\in\mathbb{Z},\ \text{age},\ \text{frames\_since\_obs}).$$

Body-frame tangent at $\mu_{bo}$: a perturbation $\xi\in\se$ acts as
$\mu' = \mu\,\Exp(\xi)$ (right-multiply). Ordering
$\xi=(\xi_\omega;\xi_v)$ — rotation first then translation. The adjoint
$$\Ad(T) = \begin{bmatrix} R & 0 \\ [t]_\times R & R \end{bmatrix}$$
moves a tangent vector through $T=(R,t)$.

\subsection*{II.2 Predict}

\textbf{World-static.} A world-static object satisfies
$T_{wb,k}\mu_{bo,k} = T_{wb,k-1}\mu_{bo,k-1}$, hence
$\mu_{bo,k} = u_k\,\mu_{bo,k-1}$ with
$u_k = T_{wb,k}^{-1}T_{wb,k-1}$. Cov:
$$\Sigma_k = \Ad(u_k)\,\Sigma_{k-1}\,\Ad(u_k)^\top + Q_{\text{static}}.$$
$Q_{\text{static}}$ schedules on \texttt{frames\_since\_obs}:
$1\!{\rm e}{-6}I_6$ (idle, $<5$), $1\!{\rm e}{-5}I_6$ ($<50$),
$1\!{\rm e}{-8}I_6$ thereafter (\file{ekf\_se3.py}).

Note the cross-coupling in $\Ad(u_k)$: the translation block of
$\Sigma$ accumulates $[t(u_k)]_\times P_{rr}[t(u_k)]_\times^\top$ each
frame, so for an unobserved track the body-frame translation
uncertainty grows in directions \emph{perpendicular to the robot's
motion}. This is correct EKF behaviour, not a bug.

\textbf{Rigid-attached.} For tracks in the held set $\mathcal{H}_k$,
$\mu_{bo,k} = \delta_{bg,k}\,\mu_{bo,k-1}$ with
$\delta_{bg,k} = T_{bg,k}T_{bg,k-1}^{-1}$. The driver applies a
phase-aware $Q_{\text{manip}}$:
\begin{itemize}
\item Seed during grasping/releasing: $\mathrm{diag}((0.02)^2{\times}3,\,(0.10)^2{\times}3)$ — physical grip-slack.
\item Seed during holding: $\mathrm{diag}(2.5\!{\rm e}{-4}{\times}3,\,1\!{\rm e}{-3}{\times}3)$ — settled grip.
\item Transitive members (apples on tray): always the holding row — they don't deform when the gripper closes around the seed.
\end{itemize}

\subsection*{II.3 Detection conditioning}

\textbf{Subpart suppression} (\file{det\_dedup.py}).
Voxelise each mask through depth at 2~cm; for ordered pair $(a,b)$
with $|V_b|\le|V_a|$, if $|V_b\cap V_a|/|V_b|>0.8$, absorb $b$ into $a$
(label histograms merge). Catches SAM2 mask splits / re-prompts.

\textbf{Centroid validity}.
$c_d^{\text{cam}} = \frac{1}{|m_d\cap\text{valid-depth}|}\sum_{p\in m_d}\mathrm{back\_project}(p,\mathrm{depth},K)$.
If the mask has no valid-depth pixels, $d$ is dropped (logged in
\texttt{centroid\_dropped}).

\subsection*{II.4 Association}

\file{pose\_update/association.py}\,$\to$\,\texttt{hungarian\_associate}.
For each $(o,d)$:
\begin{align*}
  t_{bo}^{\text{meas}} &= T_{bc}\bigl[c_d^{\text{cam}}; 1\bigr]_{1:3} \\
  R_{bo}^t &= R_{bc}\,R_{cam}\,R_{bc}^\top, \quad R_{cam} = \mathrm{diag}((0.02\,m)^2{\times}3) \\
  \nu_{od} &= t_{bo}^{\text{meas}} - \mu_{bo}[1{:}3, 4],\quad
  S_{od} = P_{tt,o} + R_{bo}^t \\
  d^2_{t,od} &= \nu_{od}^\top S_{od}^{-1} \nu_{od}.
\end{align*}
Cost matrix $C_{od} = d^2_{t,od} - \alpha\,[\tau_o = \mathrm{pid}_d] + \mathrm{label\_penalty}$ where
label\_penalty is $0$ for matching labels and $10^{12}$ otherwise
(big-M infeasibility). Hungarian minimises $\sum C_{o\sigma(o)}$;
matches with $d^2_t > G_{\text{out},t}=21.108$ are dropped.

\subsection*{II.5 Update}

\textbf{Held-track prefilters} (\file{visualize\_ekf\_tracking.py},
matched-update branch). For each accepted match $(o,d)$ with $o$ the
held seed:
\begin{itemize}
\item Held-prefilter: $\|c_d^w - (T_{wb}T_{bg})_{1:3,4}\|_2 \le 0.25$~m.
\item Innovation clamp: $\|c_d^w - (T_{wb}\mu_{bo})_{1:3,4}\|_2 \le 0.20$~m.
\end{itemize}
A failure marks the detection consumed (so it doesn't re-birth) and
skips the EKF step; the rigid-attach predict carries the track.

\textbf{Joseph step (translation-only)}, \file{gaussian\_state.py}\,$\to$
\texttt{update\_observation\_centroid}. Let $K_{tt} = P_{tt}\,S^{-1}$ ($3\times 3$):
\begin{align*}
  \delta_t^{\text{base}} &= K_{tt}\,\nu \\
  \mu_{bo}[1{:}3, 4] &\leftarrow \mu_{bo}[1{:}3, 4] + \delta_t^{\text{base}} \\
  P_{tt}^{\text{post}} &= (I - K_{tt})\,P_{tt}\,(I - K_{tt})^\top + K_{tt}\,R_{bo}^t\,K_{tt}^\top \\
  \Sigma_{bo}^{\text{post}} &= \mathrm{block\!-\!diag}\bigl(P_{tt}^{\text{post}},\ P_{rr}\bigr).
\end{align*}
The cross-covariance is zeroed: a translation-only observation
provides no information about rotation (we set $R_{rot}=\infty$),
so the honest posterior is block-diagonal. An earlier version
preserved the prior cross-cov, which broke
$P_{rr}-P_{rt}^\top P_{tt}^{-1}P_{rt}\succeq 0$ over many updates;
the refactored version is PSD by construction, with an
\texttt{eigh}-clip belt-and-suspenders.

For pairs where ICP succeeded (\file{icp\_pose.py}), an alternative
full-$\SE$ update via \texttt{joseph\_update} in
\file{object\_belief.py} replaces the centroid path.

\subsection*{II.6 Birth and prune}

Birth gates (\file{orchestrator.py}\,$\to$ \texttt{birth\_admissible}
and \file{birth\_gating.py}\,$\to$ \texttt{is\_near\_live\_track}):
\begin{enumerate}
\item Border: bbox not within \texttt{birth\_border\_margin\_px}=10~px of edge.
\item Confirm-$k$: pending-birth counter $\ge k=3$.
\item Score floor: rolling-max score $\ge 0.30$.
\item Near-live: world-frame centroid $> 0.05$~m from any same-label
  live track. For the held oid the radius switches to $0.25$~m and
  $T_{we}$ is checked as a secondary anchor.
\end{enumerate}
Survivors mint a fresh oid with $r_0 = 0.4$ and run ICP once to seed.

\textbf{Existence updates}: matched
$r' = \frac{(1-p_s)+p_s\,p_d\,\mathcal{L}\,r}{1-p_s+p_s\,r}$;
missed $r' = \frac{(1-p_d\,p_v)\,r}{1-p_d\,p_v\,r}$, $p_v$ from
ray-trace (\file{visibility.py}).

\textbf{Prune}: $r_o < 0.05$ deletes, EXCEPT the held seed which is
floored at $0.5$ (survives long held-occlusion).

\subsection*{II.7 Self-merge}

\file{orchestrator.py}\,$\to$\,\texttt{\_self\_merge\_pass}. Inputs:
held seed $s$, protected pairs
$\mathcal{P} = \{(\min,\max)\,:\,(p,c,\cdot)\in\mathcal{E}\}$.

\begin{algorithm}[H]
\caption{Self-merge pass.}
\begin{algorithmic}[1]
\For{each pair $(o_i, o_j)$ with $\ell_i = \ell_j$}
  \If{$(\min(o_i,o_j),\max(o_i,o_j))\in\mathcal{P}$} \textbf{continue} \EndIf
  \State $d \gets \|\mu_w(o_i) - \mu_w(o_j)\|_2$
  \If{$d > 0.05$} \textbf{continue} \EndIf
  \State Push $(d, d^2_t, o_i, o_j)$ as candidate.
\EndFor
\State Sort candidates by $d$ ascending.
\For{each $(d, d^2_t, o_i, o_j)$}
  \If{$o_i$ or $o_j$ already absorbed} \textbf{continue} \EndIf
  \State Choose keep$\gets s$ if $s\in\{o_i,o_j\}$, else higher-$r$, else lower oid.
  \State Information-sum-fuse $\mu, \Sigma$; merge label histograms / sam2\_tau / T\_oe / r.
  \State Append \{keep, drop, $d$, $d^2_t$\} to merges.
\EndFor
\end{algorithmic}
\end{algorithm}
The protected-pairs gate is what stops an apple-on-tray pair from
collapsing once their world centroids drift within 5~cm under
occlusion bias.

\subsection*{II.8 Relations}

\textbf{Trigger gate}
(\file{relation\_utils.py}\,$\to$\,\texttt{should\_recompute\_relations}):
fires at first call, release transitions, new-oid arrivals, or every
\texttt{relation\_every\_n\_frames}=90 frames. The grasp-transition
trigger was removed (LLM at grasp onset has incomplete information).

\textbf{Backend} (\file{relation\_orchestrator.py}\,+\,\file{relation\_client.py}).
On a triggered frame, the orchestrator builds normalised bboxes per
oid (from \texttt{det\_to\_oid}) and calls
\texttt{LLMRelationClient} or \texttt{RESTRelationClient}. The
backend returns $p\in[0,1]^{N\times N}$ with $p_{ij}$ = ``$j$ rests
on $i$''. Disk cache (\texttt{CachedRelationClient}) keyed by
$(\text{frame},N,\text{rounded bboxes})$.

\textbf{Convention swap}. Held-set expansion uses
parent=supported, child=supporter. The LLM's $(i,j)$ encodes the
opposite ($j$ rests on $i$). Edge emission swaps:
$$\mathrm{edge} = \mathrm{RelationEdge}(\text{parent}=j,\text{child}=i,\text{type}=\text{on},\text{score}=p_{ij}).$$
A regression here was responsible for held-set expansion failing —
the closure walks edges with $\text{child}\in S$, so a backward
emission means no propagation.

\textbf{EMA}
(\file{orchestrator.py}\,$\to$\,\texttt{RelationFilter}). For each
key $k=(\text{parent},\text{child},\text{type})$:
$$\mathrm{ema}_k(t) = \alpha\,\mathrm{raw}_k(t) + (1-\alpha)\,\mathrm{ema}_k(t-1),\quad \alpha=0.3.$$
Edges emitted when $\mathrm{ema}\ge 0.5$; pruned when
$\mathrm{ema}\le 0.01$. Half-life $\sim 2$ trigger calls
$\sim 180$ frames.

\textbf{Remap after merges}. When self-merge produces
$\{(\text{drop}_i,\text{keep}_i)\}$, every EMA key and emitted edge
has parent/child rewritten via the closure of \texttt{drop\_to\_keep}.
Self-loops drop, duplicate keys merge by max.

\subsection*{II.9 Held-set expansion}

\file{relation\_utils.py}\,$\to$\,\texttt{expand\_held\_with\_relations}.

\begin{algorithm}[H]
\begin{algorithmic}[1]
\State $S \gets \{s\}$
\For{$\textrm{iter}=1\ldots 8$}
  \State changed $\gets$ false
  \For{each edge $(p,c,t)\in\mathcal{E},\ t\in\{\text{in},\text{on}\}$}
    \If{$c\in S$ and $p\notin S$} $S \gets S\cup\{p\}$; changed$\gets$true \EndIf
  \EndFor
  \If{not changed} \textbf{break} \EndIf
\EndFor
\State \textbf{return} $S$
\end{algorithmic}
\end{algorithm}
The caller post-filters $S$ against currently-live oids to drop
references to pruned tracks the EMA hasn't yet decayed.

\subsection*{II.10 Gripper phase FSM}

\file{pose\_update/gripper\_state.py}\,$\to$\,\texttt{GripperPhaseTracker}.
Four phases:
$\textsc{Idle}\to\textsc{Grasping}\to\textsc{Holding}\to\textsc{Releasing}\to\textsc{Idle}$.
Driven by:
\begin{itemize}
\item Width threshold crossings with hysteresis (closed $<0.025$~m, open $\ge 0.040$~m).
\item Stability test (last $5$ widths span $<0.01$~m).
\item Min-frames timer (5) holding the transition phases through brief instability.
\end{itemize}
At the \textsc{Idle}$\to$\textsc{Grasping} edge, the held oid is
selected by \texttt{GraspOwnerDetector}: rank detections by inside-
gripper-volume containment fraction (= count\_inside / total\_pts);
tiebreak by smallest mask area. Falls back to nearest-live-track
within $0.30$~m of $T_{we}$. Self-merges propagate via
\texttt{apply\_merges}: drop $=$ \texttt{held\_obj\_id} $\to$ switch
to the keep.

\subsection*{II.11 Frame-by-frame algorithm}

\begin{algorithm}[H]
\caption{One frame of the Bernoulli-EKF tracker (driver loop).}
\begin{algorithmic}[1]
\State \emph{Inputs}: $T_{wb}$, $T_{bc}$, $T_{bg}$, $w$, RGB, depth, detections.
\State $g \gets \textsc{GripperPhaseTracker}.\textsc{step}(w, \ldots)$
\State $\textsc{relation\_pipeline}.\textsc{maybe\_update}(\ldots)$
\State $\mathcal{H} \gets \textsc{Expand}(g.\text{held},\,\mathcal{E}_t)\cap\text{live oids}$
\State \textsc{predict\_static}(skip $\mathcal{H}$);\,\textsc{rigid\_attachment\_predict}(each $o\in\mathcal{H}$)
\State \textsc{suppress\_subpart\_detections};\,\textsc{centroid\_cam\_from\_mask};\,gather \texttt{dets\_for\_assoc}
\State \textsc{hungarian\_associate} $\to$ matches, unmatched\_tracks, unmatched\_dets
\For{each match $(o, d)$}
  \If{$o = g.\text{held}$ and (held-prefilter or innov-clamp triggers)} skip update \EndIf
  \State \textsc{update\_observation\_centroid}; update $r_o$ via match log-likelihood
\EndFor
\For{each unmatched track $o$}
  \State Update $r_o$ via miss equation; floor at \texttt{r\_held\_floor} if $o = g.\text{held}$
\EndFor
\For{each unmatched detection $d$}
  \If{\textsc{birth\_admissible}$(d)$ and not \textsc{is\_near\_live\_track}$(d)$}
    \State Mint new oid, run ICP once, set $r_0=0.4$
  \EndIf
\EndFor
\State Prune oids with $r_o < r_{\min}$ (skip held seed)
\State $\mathcal{P} \gets \{(\min(p,c),\max(p,c))\,:\,(p,c,\cdot)\in\mathcal{E}_t\}$
\State \textsc{\_self\_merge\_pass}$(\text{held}=g.\text{held},\,\text{protected}=\mathcal{P})$
\State $g$.\textsc{apply\_merges}(merges); \textsc{relation\_pipeline}.\textsc{remap\_after\_merges}(merges)
\end{algorithmic}
\end{algorithm}

\subsection*{II.12 Numerical defaults summary}

\begin{tabular}{@{}lll@{}}
$R_{cam}$ & $\mathrm{diag}((0.02\,m)^2)$ & centroid measurement \\
$Q_{\text{static}}$ & $1{\rm e}{-6}\to1{\rm e}{-5}\to1{\rm e}{-8}$ & schedule on \texttt{frames\_since\_obs} \\
$Q_{\text{manip seed}}$ (graspg./relsg.) & $\mathrm{diag}((0.02)^2{\times}3,(0.10)^2{\times}3)$ & per frame \\
$Q_{\text{manip seed}}$ (holdg.) & $\mathrm{diag}(2.5{\rm e}{-4}{\times}3,1{\rm e}{-3}{\times}3)$ & per frame \\
$Q_{\text{manip transit}}$ & holding row above & apples-on-tray \\
$G_{\text{out},t}$ & $21.108$ & $\chi^2_3$ at $99.99\%$ \\
$\alpha_{\text{SAM2}}$ & $4.4$ & continuity bonus \\
\texttt{birth\_min\_dist\_m} & $0.05$ & default near-live radius \\
\texttt{held\_birth\_radius\_m} & $0.25$ & wider for held oid \\
\texttt{held\_meas\_radius\_m} & $0.25$ & $T_{we}$ prefilter \\
\texttt{held\_meas\_innov\_max\_m} & $0.20$ & innov clamp \\
\texttt{r\_min} & $0.05$ & prune threshold \\
\texttt{r\_held\_floor} & $0.50$ & held existence floor \\
\texttt{birth\_confirm\_k} & $3$ & pending-birth count \\
\texttt{birth\_score\_min} & $0.30$ & rolling-max floor \\
\texttt{self\_merge\_trans\_m} & $0.05$ & Euclidean merge gate \\
EMA $\alpha$, threshold & $0.3,\,0.5$ & relation smoothing \\
\texttt{relation\_every\_n\_frames} & $90$ & periodic LLM tick \\
\end{tabular}

\subsection*{II.13 Implementation index}
\begin{itemize}
\item State / predict: \file{pose\_update/gaussian\_state.py}, \file{pose\_update/object\_belief.py}, \file{pose\_update/ekf\_se3.py}
\item Detection conditioning: \file{pose\_update/det\_dedup.py}, \file{pose\_update/icp\_pose.py}
\item Association / update: \file{pose\_update/association.py}, \file{pose\_update/orchestrator.py}, \file{gaussian\_state.py::update\_observation\_centroid}
\item Birth / prune: \file{pose\_update/orchestrator.py::birth\_admissible}, \file{pose\_update/birth\_gating.py}
\item Self-merge: \file{pose\_update/orchestrator.py::\_self\_merge\_pass}
\item Relations: \file{pose\_update/relation\_orchestrator.py}, \file{pose\_update/relation\_client.py}, \file{pose\_update/relation\_utils.py}
\item Gripper FSM: \file{pose\_update/gripper\_state.py}, \file{pose\_update/grasp\_owner\_detector.py}
\item Visibility: \file{pose\_update/visibility.py}
\item Driver / harness: \file{tests/visualize\_ekf\_tracking.py}
\end{itemize}

\end{document}
